---
title: "# Reproducible Research: Peer Assessment 1"
author: "Shumann"
date: "Saturday, August 16, 2014"
output: html_document
---
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self"" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.




```{r echo=TRUE, echo=TRUE}

library(caret)
library(reshape)

```

## Loading and preprocessing the data

1. The required libraries were loaded into R (analysis interface).

2. The activity monitoring data was read into a data frame for further processing.  The date field was read as strings instead of factors(default).

3. All records with steps = "NA"were removed. Exploring data showed 2304 records with steps = "NA"

```{r, echo=TRUE, eval=TRUE}

# unzip("repdata-Fdata-Factivity.zip")
actData <- read.csv("activity.csv", header=T, stringsAsFactors=F)
# actData <- actData[!is.na(actData$steps)==T, ]

```

## What is mean total number of steps taken per day?

```{r, echo=TRUE, eval=TRUE}

# Find mean with tapply and melt (from reshape)
totalStep <- with(actData, tapply(steps, date, sum, na.rm=T));
totalStep <- melt(totalStep);

# Add column names
colnames(totalStep) <- c("date", "step");
totalStep$date <- as.Date(as.character(totalStep$date))

# Print Histogram with Mean and Median
histStep <- with(totalStep, hist(step, breaks=5,  col="grey", xlab="Step", ylab="Frequency", main="Histogram- Total Number of Steps Each Day"))

# Add Mean and Median to the Histogram
abline(v=mean(totalStep$step), lwd=3, lty="dashed", col="blue") # Mean - Blue Line
abline(v=median(totalStep$step), lwd=3, lty="dashed", col="red") # Median Red line

# Print Mean and Median of Total number of steps taken each day
print(mean(totalStep$step))
print(median(totalStep$step))
```


## What is the average daily activity pattern?

```{r, echo=TRUE, eval=TRUE}

# Find mean with tapply and melt (from reshape)
avgStep <- with(actData, tapply(steps, interval, mean, na.rm=T));
avgStep <- melt(avgStep);

# Add column names
colnames(avgStep) <- c("interval", "step");

# Print Histogram with Mean and Median
tsStep <- with(avgStep, xyplot(step ~ interval, type="l", xlab="interval", ylab="Average Step", main="Average Step per Interval"))
print(tsStep);

# Find the record with max step(averaged across days)
IntMaxStep <- avgStep[avgStep$step==max(avgStep$step), ]
print(IntMaxStep)
```


## Imputing missing values

```{r, echo=TRUE, eval=TRUE}
# Total Missing values in "steps"
totalNA <- (actData[is.na(actData$step)==T, ])

# Total records with "NA" (rows)
length(totalNA$steps)

# replace "NA" with mean for the intervals across all dates

library(plyr)

impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
newData <- ddply(actData, ~ steps, transform, steps = impute.mean(steps))
# newData[order(newData$date), ]

# Find mean with tapply and melt (from reshape)
totalStepnew <- with(newData, tapply(steps, date, sum, na.rm=T));
totalStepnew <- melt(totalStepnew);

# Add column names
colnames(totalStepnew) <- c("date", "step");
totalStepnew$date <- as.Date(as.character(totalStepnew$date))

# Print Histogram with Mean and Median
histStep <- with(totalStepnew, hist(step, breaks=5,  col="grey", xlab="Step", ylab="Frequency", main="Histogram- Total Number of Steps Each Day"))

# Add Mean and Median to the Histogram
abline(v=mean(totalStepnew$step), lwd=3, lty="dashed", col="blue") # Mean - Blue Line
abline(v=median(totalStepnew$step), lwd=3, lty="dashed", col="red") # Median Red line

# Print Mean and Median of Total number of steps taken each day
print(mean(totalStepnew$step))
print(median(totalStepnew$step))

```


## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE, eval=TRUE}
# Create factor variabe


#totalNA <- (actData[is.na(actData$step)==T, ]) add.S

